---
title: 'STAT5003 Group Project — Final Report'
subtitle: 'Predicting Zoned Energy Star Ratings from Technical Specifications'
author: "Mahek Poddar, Kashyap Adipudi, Atharva Gadgil, Shreyash Wetal, Kavvin Udhayakumar Valarmathi (W02G10)"
date: "2026-04-17"
output:
  html_document:
    code_folding: hide
    theme: flatly
    toc: true
    toc_float: true
knit: (function(input, ...) rmarkdown::render(input, output_format = "html_document"))
---

```{css, echo=FALSE}
/* CSS to enforce exactly a 6-page layout when printed to PDF */
body { font-size: 14px; line-height: 1.6; }
/* Force major sections to start on a new page */
h1 { font-size: 22px; margin-top: 20px; margin-bottom: 15px; page-break-before: always; color: #2C3E50; border-bottom: 2px solid #ECF0F1; padding-bottom: 5px;}
h1:first-of-type { page-break-before: auto; } /* Keep first header on page 1 */
h2 { font-size: 18px; margin-top: 15px; margin-bottom: 10px; color: #2980B9; }
h3 { font-size: 15px; font-weight: bold; margin-top: 12px; }
.table { margin-bottom: 15px; font-size: 12px; }
.caption { color: #333; font-weight: bold; padding-bottom: 5px; font-size: 13px; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)
suppressPackageStartupMessages({
  # Load MASS first so dplyr::select, dplyr::filter overwrite it after
  library(MASS)
  library(tidyverse); library(ggplot2); library(corrplot)
  library(scales);    library(knitr);   library(kableExtra)
  library(patchwork); library(forcats); library(gridExtra)
  library(caret);     library(nnet)
  library(randomForest); library(xgboost); library(class)
})
# Explicitly ensure dplyr functions are used (MASS masks select/filter)
select <- dplyr::select
filter <- dplyr::filter
set.seed(42)
```

# 1. Framing & Dataset Selection

**Research Question:** Can an air conditioner's Zoned Energy Star Rating for mixed climates (`c_star_mixed`) be accurately predicted using pre-rating technical specifications and performance measurements, enabling automated efficiency classification without requiring a formal ZERL test submission?

Australia's residential and commercial cooling industry uses about 40% of household power during summer peaks (DCCEEW, 2023). In order to more precisely identify efficiency in the three temperature zones (cold, mixed, and hot), the Australian government updated the antiquated 2010 label with the Zoned Energy Rating Label (ZERL) under the GEMS Determination 2019 (Commonwealth of Australia, 2019). Currently, obtaining a star rating requires costly, approved laboratory testing that takes weeks and costs tens of thousands of dollars for each product variation.

Our investigation is relevant to three groups:
1. **Consumers** benefit from being able to estimate likely star performance from publicly available specifications before purchase. 
2. **Regulators** (E3 Programme) can use a trained model to flag potentially mislabelled registrations for audit. 
3. **Manufacturers** can estimate expected ratings during R&D, before committing to a formal test submission and accelerating their development cycle. 

This is a **multi-class ordinal classification problem** — the target variable takes 15 discrete ordered values from 0 to 7 in 0.5-step increments.

---

# 2. Consider & Gather Data

**Registry:** Australian Government Energy Rating Register — https://www.energyrating.gov.au/  
**File:** `ac_2026_04_10.csv` (31 March 2026) | **Licence:** Creative Commons Attribution 3.0 Australia  
**Regulatory basis:** E3 Programme, EE Regulations 2020 Sch. 2A, GEMS AC Determination 2019

```{r load-data, results='hide'}
# Looks for the CSV in the same folder as this Rmd file
csv_path <- file.path(dirname(rstudioapi::getSourceEditorContext()$path),
                      "ac_2026_04_10.csv")
# Fallback: if not in RStudio, use working directory
if (!file.exists(csv_path)) csv_path <- "ac_2026_04_10.csv"
df_raw <- read_csv(csv_path, show_col_types = FALSE)
```

Our dataset satisfies **four** of the five rubric criteria, visualised in **Figure 1**:

**1. Messy:** The dataset contains substantial structured missingness. 10 columns are entirely empty and carry zero information. The structured missingness is informative as it reflects which optional test protocols manufacturers chose to complete, which we plan to capture via binary indicator variables during preprocessing.
**2. Complex:** With 167 features spanning continuous performance measurements, ordinal star ratings, nominal categories, binary flags, and high-cardinality identifiers (127 brands, 5,534 unique model numbers), the dataset presents substantial dimensionality challenges. 
**3. Multi-class:** The target variable has 15 ordered classes, making this a genuine ordinal multi-class problem. Very low ratings (<2.0) are increasingly rare, while stars 6–7 have only 35 combined units, requiring class-weighting.
**4. Computationally Demanding:** The combination of 167 characteristics, 15 ordered target classes, and the requirement for stratified cross-validation make our dataset computationally challenging.

```{r fig1-dashboard, fig.cap="**Figure 1:** Dataset Justification Dashboard. Shows missingness profile, multi-class target distribution, and perfect data leakage bands from regulatory formulas.", fig.height=5.5, fig.width=10}
# Clean Data first for the plots
cols_empty <- c("avg_pwr_standby_mode","Setting_heat","eermepslev","TestedOutputEER",
                "TestedOutputCOP","PL_EERMEPS","PL_COPMEPS","GrandDate",
                "star2000_cool","star2000_heat")
df <- df_raw %>% select(-all_of(cols_empty))
char_to_num <- c("EERtestAvg","COPtestAvg","H-Power_Inp_Rated","H-Total Heat Rated",
                 "C-Power_Inp_Rated","C-Total Cool Rated","Rated AEER","Rated ACOP",
                 "c_star_cold","c_star_mixed","c_star_hot")
df <- df %>%
  mutate(across(all_of(intersect(char_to_num, names(df))), ~ suppressWarnings(as.numeric(.))),
         h_star_cold  = suppressWarnings(as.numeric(na_if(as.character(h_star_cold),  "-"))),
         h_star_mixed = suppressWarnings(as.numeric(na_if(as.character(h_star_mixed), "-"))),
         Calculated_EER = `C-Total Cool Rated` / `C-Power_Inp_Rated`)

# 1A: Missingness
miss_df <- df_raw %>% summarise(across(everything(), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "pct") %>% filter(pct > 0)
p1a <- miss_df %>% arrange(desc(pct)) %>% slice_head(n = 12) %>%
  ggplot(aes(x = fct_reorder(var, pct), y = pct)) + geom_col(fill="#E67E22") +
  coord_flip() + theme_minimal(base_size=10) + labs(title="1A: Top Missing Columns", x=NULL, y="% Missing")

# 1B: Target Distribution
p1b <- df %>% filter(!is.na(c_star_mixed)) %>% count(c_star_mixed) %>%
  ggplot(aes(x = factor(c_star_mixed), y = n, fill = c_star_mixed == 0)) +
  geom_col(show.legend=FALSE) + scale_fill_manual(values=c("TRUE"="#C0392B","FALSE"="#2980B9")) +
  theme_minimal(base_size=10) + labs(title="1B: 15-Class Target Distribution", x="Star Rating (c_star_mixed)", y="Count")

# 1C: Leakage
t_check <- df %>% filter(!is.na(c_star_cold), c_star_cold > 0) %>%
  mutate(TCSPF = as.numeric(`Residential TCSPF_cold`)) %>% group_by(c_star_cold) %>%
  summarise(T_min=min(TCSPF,na.rm=TRUE), T_max=max(TCSPF,na.rm=TRUE))
p1c <- t_check %>% ggplot(aes(x = factor(c_star_cold))) +
  geom_linerange(aes(ymin=T_min, ymax=T_max), linewidth=4, colour="#8E44AD", alpha=0.8) +
  theme_minimal(base_size=10) + labs(title="1C: TCSPF Leakage (Perfect Bands)", subtitle="Each star occupies a 0.5-wide interval", x="Star Class", y="TCSPF Range")

(p1a | p1b) / p1c
```

---

# 3. Data Cleaning & Target Leakage

We must find characteristics that **define** the star rating instead of predicting it before performing any correlation study. The regulation formula input from which stars are directly calculated is the TCSPF (Total Cooling Season Performance Factor). As shown in **Figure 1C**, every star class corresponds to a single 0.5-wide TCSPF band. 

The Spearman correlation between TCSPF and the star rating is 0.99. All TCSPF, tcec, thec, and HSPF columns are therefore **excluded** from all modelling — they are outputs of the same regulatory formula that produces the star, not independent predictors.

### 3.1 EER Predictor Comparison

```{r eer-comparison}
tibble(
  Predictor      = c("Rated AEER", "EERtestAvg", "Calculated_EER (naive)"),
  `Pearson r`    = c(round(cor(df$`Rated AEER`,  df$c_star_mixed, use="pairwise.complete.obs"), 3),
                     round(cor(df$EERtestAvg,     df$c_star_mixed, use="pairwise.complete.obs"), 3),
                     round(cor(df$Calculated_EER, df$c_star_mixed, use="pairwise.complete.obs"), 3)),
  `Spearman rho` = c(round(cor(df$`Rated AEER`,  df$c_star_mixed, use="pairwise.complete.obs", method="spearman"), 3),
                     round(cor(df$EERtestAvg,     df$c_star_mixed, use="pairwise.complete.obs", method="spearman"), 3),
                     round(cor(df$Calculated_EER, df$c_star_mixed, use="pairwise.complete.obs", method="spearman"), 3)),
  Issue = c("None — preferred predictor", "None — good alternative", "One outlier (EER=7,937) destroys Pearson r")
) %>% kable(caption = "EER predictor comparison — Rated AEER is the correct choice") %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE, font_size = 12)
```

A naive `Calculated_EER = Cooling Capacity / Power Input` produces one extreme outlier (EER = 7,937) from a row with near-zero power input, which collapses the Pearson correlation to 0.014 even though the true relationship is strong. `Rated AEER` avoids this issue entirely and will be used as the primary EER predictor.

### 3.2 Target Selection

```{r target-selection}
target_candidates <- c("c_star_cold","c_star_mixed","c_star_hot",
                       "h_star_cold","h_star_mixed","h_star_hot")

# Coerce all needed columns to numeric before correlation
extra_num <- c("Rated AEER","Rated ACOP","EERtestAvg","COPtestAvg",
               "C-Total Cool Rated","C-Power_Inp_Rated",
               "indoor_sound_level","outdoor_sound_level")
all_to_coerce <- c(target_candidates, extra_num)
for (v in intersect(all_to_coerce, names(df)))
  df[[v]] <- suppressWarnings(as.numeric(df[[v]]))

legit_num <- c("EERtestAvg","COPtestAvg","Rated AEER","Rated ACOP",
               "C-Total Cool Rated","C-Power_Inp_Rated",
               "indoor_sound_level","outdoor_sound_level")
legit_cat <- c("Refrigerant","Configuration2","Configuration1","Type","Phase")

avg_r <- sapply(target_candidates, function(tgt) {
  mean(sapply(legit_num, function(feat) {
    sub <- df %>% select(all_of(c(feat,tgt))) %>% drop_na()
    if (nrow(sub)>100) abs(cor(sub[[feat]],sub[[tgt]])) else NA_real_
  }), na.rm=TRUE)
})

compute_eta2 <- function(tgt, grp, data) {
  sub <- data %>% select(all_of(c(tgt,grp))) %>% drop_na()
  groups <- split(sub[[tgt]], sub[[grp]]); groups <- groups[sapply(groups,length)>=5]
  if (length(groups)<2) return(NA_real_)
  all_v <- unlist(groups); gm <- mean(all_v)
  sum(sapply(groups, function(g) length(g)*(mean(g)-gm)^2)) / sum((all_v-gm)^2)
}

avg_eta2 <- sapply(target_candidates, function(tgt)
  mean(sapply(legit_cat, function(cat) compute_eta2(tgt,cat,df)), na.rm=TRUE))

target_scores <- tibble(
  Target    = target_candidates,
  Coverage  = paste0(sapply(target_candidates, function(v) round(mean(!is.na(df[[v]]))*100,1)),"%"),
  Classes   = sapply(target_candidates, function(v) n_distinct(df[[v]], na.rm=TRUE)),
  `avg|r|`  = round(avg_r,3), `eta²` = round(avg_eta2,3),
  Composite = round((avg_r+avg_eta2)/2,3)) %>% arrange(desc(Composite))

target_scores %>%
  kable(caption="Target selection: composite predictability score (legitimate predictors only)") %>%
  kable_styling(bootstrap_options=c("striped"), full_width=FALSE, font_size=12) %>%
  row_spec(which(target_scores$Target=="c_star_mixed"), bold=TRUE, background="#EAF3DE")
```

**Decision:** We selected `c_star_mixed` as our primary target. With the highest composite predictability score (0.413) across categorical and numeric variables, it is the most learnable target from legitimate pre-rating features. Its 98.9% data coverage ensures models are trained on nearly the full dataset rather than a biased subset, and the mixed-climate zone is the rating consumers and regulators see on the label (Commonwealth of Australia, 2019).

---

# 4. Exploratory Data Analysis (EDA)

```{r fig2-dashboard, fig.cap="**Figure 2:** Univariate distributions, key categorical variables, and correlations.", fig.height=7.5, fig.width=10}
# 2A: AEER Histogram
p2a <- df %>% filter(!is.na(`Rated AEER`)) %>% ggplot(aes(x = `Rated AEER`)) +
  geom_histogram(bins=35, fill="#27AE60", colour="white") + theme_minimal(base_size=10) +
  labs(title="2A: Rated AEER Distribution", x="AEER", y="Count")

# 2B: Capacity Histogram
p2b <- df %>% filter(!is.na(`C-Total Cool Rated`)) %>% ggplot(aes(x = `C-Total Cool Rated`)) +
  geom_histogram(bins=40, fill="#2980B9", colour="white") + scale_x_log10(labels=comma) +
  theme_minimal(base_size=10) + labs(title="2B: Cooling Capacity (log)", x="kW (log10)", y="Count")

# 2C: Refrigerants
p2c <- df %>% count(Refrigerant) %>% slice_max(n, n=4) %>%
  ggplot(aes(x=fct_reorder(Refrigerant, n), y=n)) + geom_col(fill="#8E44AD") + coord_flip() +
  theme_minimal(base_size=10) + labs(title="2C: Top Refrigerants", x=NULL, y="Count")

# 2D: Configurations
p2d <- df %>% count(Configuration2) %>% slice_max(n, n=4) %>%
  ggplot(aes(x=fct_reorder(Configuration2, n), y=n)) + geom_col(fill="#D35400") + coord_flip() +
  theme_minimal(base_size=10) + labs(title="2D: Top Configurations", x=NULL, y="Count")

((p2a | p2b) / (p2c | p2d))
```

**Univariate Insights:** Figure 2A shows that efficiency metrics are approximately normal, bounded by MEPS floors. Cooling capacity (Figure 2B) is right-skewed; residential 2–20 kW dominates, with a commercial tail to 232 kW requiring log transformation. Single-split systems dominate configuration, R32 and R410A are the primary refrigerants.

```{r corrplot, fig.height=4, fig.width=9}
corr_vars <- c("Rated AEER","Rated ACOP","EERtestAvg","COPtestAvg",
               "C-Total Cool Rated","C-Power_Inp_Rated",
               "indoor_sound_level","outdoor_sound_level","c_star_mixed")
corr_data <- df %>% select(all_of(corr_vars)) %>% drop_na()
corrplot(cor(corr_data, use="pairwise.complete.obs"),
         method="color", type="upper", tl.cex=0.7, tl.col="black",
         addCoef.col="black", number.cex=0.55,
         col=colorRampPalette(c("#C0392B","white","#27AE60"))(200),
         title="Figure 2E: Correlation matrix — key numeric features", mar=c(0,0,1.5,0))
```

AEER and ACOP shows the strongest positive correlation with star rating. Sound levels are showing moderate negative correlation, confirming quieter units tend to score higher stars.

---

# 4. Exploratory Data Analysis (Continued)

```{r fig3-bivariate, fig.cap="**Figure 3:** Multivariate interactions. Note the density overlap of classes and the impact of sound levels.", fig.height=6, fig.width=10}
# 3A: Boxplot
p3a <- df %>% filter(!is.na(c_star_mixed), !is.na(`Rated AEER`)) %>%
  ggplot(aes(x = factor(c_star_mixed), y = `Rated AEER`, fill = factor(c_star_mixed))) +
  geom_boxplot(outlier.alpha = 0.3, show.legend = FALSE) + scale_fill_viridis_d() +
  theme_minimal(base_size=10) + labs(title="3A: AEER vs Star Class", x="Star Rating", y="Rated AEER")

# 3B: Heatmap Interaction
p3b <- df %>% filter(!is.na(c_star_mixed), Refrigerant %in% c("R32","R410A","R290"), Configuration2 %in% c("Single Split System","Portable","Packaged","vrf")) %>%
  group_by(Refrigerant, Configuration2) %>% summarise(m_star = mean(c_star_mixed, na.rm=TRUE), .groups="drop") %>%
  ggplot(aes(x=Configuration2, y=Refrigerant, fill=m_star)) + geom_tile(colour="white") +
  geom_text(aes(label=round(m_star,1)), size=3) + scale_fill_gradient2(low="#C0392B", mid="#F39C12", high="#27AE60", midpoint=2.5) +
  theme_minimal(base_size=10) + theme(axis.text.x = element_text(angle=15, hjust=1), legend.position="none") +
  labs(title="3B: Ref × Config Interaction", x=NULL, y=NULL)

# 3C: Sound Outdoor
p3c <- df %>% filter(!is.na(c_star_mixed), !is.na(outdoor_sound_level)) %>%
  ggplot(aes(x = outdoor_sound_level, y = c_star_mixed)) +
  geom_point(alpha = 0.2, size = 1, colour = "#16A085") +
  geom_smooth(method = "loess", colour = "red", linewidth = 1, se = FALSE) +
  theme_minimal(base_size=10) + labs(title="3C: Outdoor Sound vs Star", subtitle="Proxy for inverter technology", x="Outdoor Sound (dB)", y="c_star_mixed")

(p3a | p3b) / p3c
```

**Bivariate & Interactions:** Figure 3A confirms a clear monotonic relationship between AEER and star rating — median AEER increases approximately 0.15 units per half-star. However, adjacent classes heavily overlap, confirming no single predictor can cleanly separate all 15 classes.

Figure 3B reveals a critical interaction: R32 refrigerant units consistently outperform R410A units by approximately one full star, but R290 units (restricted to Portables) are anchored at 0 stars by regulation. Figure 3C shows that quieter units tend to score higher stars — sound level serves as a useful proxy for inverter technology.

### 4.1 Categorical Hypothesis Testing & Data Quality

```{r kruskal-wallis}
kw_config <- kruskal.test(c_star_mixed ~ Configuration2, data = df)
kw_refrig <- kruskal.test(c_star_mixed ~ Refrigerant,    data = df)

tibble(
  `Categorical Predictor` = c("Configuration2", "Refrigerant"),
  `Kruskal-Wallis H` = c(round(kw_config$statistic, 1), round(kw_refrig$statistic, 1)),
  `df` = c(kw_config$parameter, kw_refrig$parameter),
  `p-value` = c(format.pval(kw_config$p.value, digits=3), format.pval(kw_refrig$p.value, digits=3)),
  `Conclusion` = c("Reject H0: Stars differ significantly", "Reject H0: Stars differ significantly")
) %>% kable(caption="Kruskal-Wallis Rank Sum Tests (Used due to sparse contingency cells)") %>% 
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE, font_size = 12)
```

**Data Quality Insight (The Portable Paradox):** Despite having legitimate and at times high AEER values, 100% of portable air conditioners are given zero stars. This regulation pertains to Product Class 4. To enable the model to learn this stringent regulatory override, we must create a binary flag called `is_portable`.

---

# 5. Project Planning & Methodology

Based on the insights derived from our EDA, we have structured a robust pipeline.

### 5.1 Preprocessing Pipeline

```{r preprocessing-plan}
tibble(
  Step = 1:5,
  Action = c(
    "Drop completely empty & >85% missing columns",
    "Exclude leaky features (TCSPF, tcec, thec, HSPF)",
    "Median imputation for remaining numeric predictors",
    "One-hot encode Refrigerant, Configuration, Type",
    "StandardScaler for linear/distance-based models"
  ),
  Justification = c(
    "Too sparse for reliable imputation; removes noise.",
    "TCSPF band width = 0.5 per star — defines the target perfectly.",
    "Preserves all rows; robust to right skew in capacity features.",
    "Required for Logistic Regression and KNN algorithms.",
    "Tree models are scale-invariant; distance models require it."
  )
) %>% kable(caption="Preprocessing Plan") %>% 
  kable_styling(bootstrap_options = c("striped"), font_size = 12)
```

### 5.2 Models & Evaluation Metrics

```{r model-plan}
tibble(
  Model = c("Multinomial Logistic Regression", "Ordinal Logistic Regression (polr)", "Random Forest", "Gradient Boosted Trees (XGBoost)", "k-Nearest Neighbours"),
  Justification = c(
    "Linear baseline; interpretable coefficients; establishes minimum accuracy floor.",
    "Respects ordinal structure; calibrated cumulative probabilities; most theoretically appropriate.",
    "Handles mixed feature types natively; extracts feature importance; robust to outliers.",
    "Best out-of-box accuracy on tabular data; captures AEER × Configuration interactions.",
    "Non-parametric; validates whether feature space forms coherent spatial clusters."
  )
) %>% kable(caption="Five classification models with justification") %>% 
  kable_styling(bootstrap_options = c("striped"), font_size = 12)
```

**Evaluation Strategy:** 
We will be evaluating the models using **Overall accuracy** (primary reporting metric), **Weighted F1-score** (which accounts for class imbalance by weighting each class by its sample frequency), and **Ordinal Mean Absolute Error (MAE)** (penalises large mis-predictions proportionally). 

### 5.3 Timeline & Responsibilities

```{r timeline}
tibble(
  Week = c("Week 8", "Week 9", "Week 10", "Week 11", "Week 12"),
  Milestone = c(
    "Preprocessing pipeline; baseline models (LR, Ordinal LR)",
    "Random Forest; feature importance analysis",
    "XGBoost; hyperparameter tuning via cross-validation",
    "KNN; full model comparison table compiled",
    "Draft final report written; presentation rehearsed"
  ),
  Responsible = c("Mahek, Kashyap", "Atharva, Shreyash", "Mahek, Kashyap, Atharva", "Shreyash, Kavvin", "All Members")
) %>% kable(caption="Project timeline with team responsibilities") %>% 
  kable_styling(bootstrap_options = c("striped"), font_size = 12)
```

---

# 6. References & Declarations

Breiman, L. (2001). Random forests. *Machine Learning, 45*(1), 5–32. https://doi.org/10.1023/A:1010933404324

Chen, T., & Guestrin, C. (2016). XGBoost: A scalable tree boosting system. *Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining*, 785–794. https://doi.org/10.1145/2939672.2939785

Commonwealth of Australia. (2019). *Greenhouse and Energy Minimum Standards (Air Conditioners up to 65kW) Determination 2019*. Federal Register of Legislation. https://www.legislation.gov.au/Details/F2019L00490

Department of Climate Change, Energy, the Environment and Water. (2023). *Energy Rating Data for Household Appliances – Air Conditioners* [Dataset]. Australian Government. https://www.energyrating.gov.au/

Hollander, M., Wolfe, D. A., & Chicken, E. (2013). *Nonparametric statistical methods* (3rd ed.). Wiley.

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning with applications in R* (2nd ed.). Springer. https://doi.org/10.1007/978-1-0716-1418-1

Kuhn, M., & Johnson, K. (2013). *Applied predictive modeling*. Springer. https://doi.org/10.1007/978-1-4614-6849-3

Standards Australia. (2014). *AS/NZS 3823.4.1:2014 — Performance of electrical appliances: Air conditioners and heat pumps — Part 4.1: Zoned energy rating*. Standards Australia.

Standards Australia. (2012). *AS/NZS 3823.1.1:2012 — Performance of electrical appliances: Air conditioners and heat pumps*. Standards Australia.

### AI Use Statement

We acknowledge the use of Claude Sonnet 4.6 (Anthropic, https://claude.ai) to assist with reviewing code chunks for robustness, restructuring the  grids using the `patchwork` library, and resolving CSS layout issues for strict PDF pagination constraints. All original data analysis, interpretation of statistical results (Kruskal-Wallis), model selection decisions, feature engineering and final conclusions were determined independently by us.
---

# 3. Model Implementation

## 3.1 Preprocessing & Feature Engineering

All preprocessing steps follows  directly from the EDA findings: leaky TCSPF columns are excluded (Spearman r = 0.99), cooling capacity is log-transformed due to right skew (Figure 2B), and the portable paradox necessitates a dedicated binary flag.

```{r preprocessing}
# Step 1 & 2: Drop leaky and empty columns
leaky_cols <- names(df)[grepl("TCSPF|tcec|thec|HSPF|Residential|Commercial",
                               names(df), ignore.case = TRUE)]

df_model <- df %>%
  select(-all_of(intersect(leaky_cols, names(df)))) %>%
  filter(!is.na(c_star_mixed)) %>%
  # Step 3: Feature engineering from EDA insights
  mutate(
    is_portable  = as.integer(Configuration2 == "Portable"),
    is_R32       = as.integer(Refrigerant    == "R32"),
    is_split     = as.integer(Configuration2 == "Single Split System"),
    log_capacity = log1p(`C-Total Cool Rated`)
  ) %>%
  # Step 4: Select final feature set
  select(
    c_star_mixed,
    `Rated AEER`, `Rated ACOP`,
    outdoor_sound_level, indoor_sound_level,
    log_capacity,
    is_portable, is_R32, is_split
  ) %>%
  # Step 5: Median-impute numeric predictors (matches Section 5.1 plan).
  # Only drop rows still missing the core efficiency or capacity features,
  # which are the actual predictors of the star rating.
  mutate(
    outdoor_sound_level = ifelse(is.na(outdoor_sound_level),
                                 median(outdoor_sound_level, na.rm = TRUE),
                                 outdoor_sound_level),
    indoor_sound_level  = ifelse(is.na(indoor_sound_level),
                                 median(indoor_sound_level,  na.rm = TRUE),
                                 indoor_sound_level)
  ) %>%
  filter(!is.na(`Rated AEER`),
         !is.na(`Rated ACOP`),
         !is.na(log_capacity))

cat("Rows available for modelling:", nrow(df_model), "\n")
cat("Features used:", ncol(df_model) - 1, "\n")
cat("Target classes:", n_distinct(df_model$c_star_mixed), "\n")
cat("Class distribution:\n")
print(table(df_model$c_star_mixed))
```

```{r train-test-split}
# ── Step 5: Stratified 80/20 train-test split ─────────────────────────────────
set.seed(42)
train_idx <- createDataPartition(df_model$c_star_mixed, p = 0.8, list = FALSE)
train_df  <- df_model[ train_idx, ]
test_df   <- df_model[-train_idx, ]

# Separate features and target
X_train <- train_df %>% select(-c_star_mixed)
X_test  <- test_df  %>% select(-c_star_mixed)
y_train <- train_df$c_star_mixed
y_test  <- test_df$c_star_mixed

# ── Step 6: Scale for LR and KNN (tree models do not need scaling) ────────────
X_train_sc <- scale(X_train)
X_test_sc  <- scale(X_test,
                    center = attr(X_train_sc, "scaled:center"),
                    scale  = attr(X_train_sc, "scaled:scale"))

cat("Train rows:", nrow(train_df), "| Test rows:", nrow(test_df), "\n")
```

```{r preprocessing-table, echo=FALSE}
tibble(
  Step = 1:7,
  Action = c(
    "Drop completely empty & >85% missing columns",
    "Exclude leaky features (TCSPF, tcec, thec, HSPF)",
    "Log-transform cooling capacity",
    "Engineer is_portable binary flag",
    "Engineer is_R32 and is_split binary flags",
    "Drop remaining NA rows after feature selection",
    "StandardScaler applied for LR and KNN only"
  ),
  Justification = c(
    "Zero-information columns removed to reduce noise.",
    "TCSPF band width = 0.5 per star — directly defines the target (Spearman r = 0.99).",
    "Right-skewed distribution confirmed in EDA Figure 2B; log normalises.",
    "Portable paradox: regulatory override always yields 0 stars regardless of AEER.",
    "EDA (Figure 3B) showed R32 and split systems significantly outperform others.",
    "drop_na() retains maximum rows; feature set has low missingness after selection.",
    "Tree models are scale-invariant; distance and linear models require normalisation."
  )
) %>%
  kable(caption = "Full preprocessing and feature engineering pipeline") %>%
  kable_styling(bootstrap_options = c("striped"), font_size = 12, full_width = TRUE)
```

## 3.2 Evaluation Metric Helper

All five models are evaluated consistently using the same helper function below, ensuring fair comparison across accuracy, weighted F1-score, and ordinal MAE.

```{r metrics-helper}
# ── Consistent evaluation function used for all 5 models ─────────────────────
get_metrics <- function(pred, actual) {
  pred_num   <- as.numeric(as.character(pred))
  actual_num <- as.numeric(as.character(actual))
  all_levels <- sort(unique(actual_num))

  pred_f   <- factor(pred_num,   levels = all_levels)
  actual_f <- factor(actual_num, levels = all_levels)

  # Overall accuracy
  acc <- mean(pred_num == actual_num)

  # Ordinal MAE — penalises large misclassifications proportionally
  mae <- mean(abs(pred_num - actual_num))

  # Weighted F1 — accounts for class imbalance
  cm <- confusionMatrix(pred_f, actual_f)
  sens <- cm$byClass[, "Sensitivity"]
  ppv  <- cm$byClass[, "Pos Pred Value"]
  f1   <- 2 * sens * ppv / (sens + ppv + 1e-10)
  support <- as.numeric(table(actual_f))
  wf1  <- sum(f1 * support / sum(support), na.rm = TRUE)

  c(Accuracy = round(acc, 4),
    `Weighted F1` = round(wf1, 4),
    MAE = round(mae, 4))
}
```

## 3.3 Multinomial Logistic Regression (Baseline)

Multinomial LR is fitted first as a linear baseline. It establishes the minimum accuracy floor and provides interpretable coefficients to validate feature directions against EDA findings.

```{r model-mlr}
set.seed(42)
mlr_fit <- multinom(
  c_star_mixed ~ .,
  data    = data.frame(c_star_mixed = y_train, X_train_sc),
  MaxNWts = 5000, maxit = 300, trace = FALSE
)
mlr_pred <- predict(mlr_fit, newdata = data.frame(X_test_sc))
mlr_m    <- get_metrics(mlr_pred, y_test)
cat("MLR  | Accuracy:", mlr_m["Accuracy"],
    "| Weighted F1:", mlr_m["Weighted F1"],
    "| MAE:", mlr_m["MAE"], "\n")
```

## 3.4 Ordinal Logistic Regression (polr)

Ordinal LR explicitly respects the ordered structure of `c_star_mixed`, modelling cumulative probabilities rather than treating classes as independent. This is the most theoretically appropriate model for an ordered outcome.

```{r model-olr}
set.seed(42)
olr_fit <- polr(
  factor(c_star_mixed, ordered = TRUE) ~ .,
  data   = data.frame(c_star_mixed = y_train, X_train_sc),
  Hess   = TRUE, method = "logistic"
)
olr_pred <- as.numeric(as.character(
  predict(olr_fit, newdata = data.frame(X_test_sc))
))
olr_m <- get_metrics(olr_pred, y_test)
cat("OLR  | Accuracy:", olr_m["Accuracy"],
    "| Weighted F1:", olr_m["Weighted F1"],
    "| MAE:", olr_m["MAE"], "\n")
```

## 3.5 Random Forest

```{r model-rf}
# set.seed(42)
# rf_fit  <- randomForest(x = X_train, y = factor(y_train), ntree = 300, mtry = 3, importance = TRUE)
# rf_pred <- as.numeric(as.character(predict(rf_fit, newdata = X_test)))
# rf_m    <- get_metrics(rf_pred, y_test)
# cat("RF   | Accuracy:", rf_m["Accuracy"], "| Weighted F1:", rf_m["Weighted F1"], "| MAE:", rf_m["MAE"], "\n")

# Placeholder so report knits until section is complete
rf_pred <- rep(median(y_test), length(y_test))
rf_m    <- get_metrics(rf_pred, y_test)
```

## 3.6 XGBoost

XGBoost (Chen & Guestrin, 2016) builds an additive ensemble of decision trees, where each tree corrects the residual errors of its predecessors. Key advantages for this problem:

- Captures non-linear interactions (e.g. AEER × refrigerant type) identified in EDA.
- Natively handles mixed numeric and binary features without requiring feature scaling.
- Built-in regularisation (`subsample`, `colsample_bytree`) reduces overfitting on the sparse high-star classes.
- Provides interpretable feature importance scores via information gain.

The `multi:softmax` objective is used for the 15-class target. XGBoost requires 0-indexed integer labels, so `c_star_mixed` values are re-encoded before training and mapped back afterwards.

### 3.6.1 Naive Baseline

Before fitting XGBoost, a **majority-class baseline** is established. This predicts the most frequent training class for every test observation — the minimum bar any model must beat to demonstrate genuine learning.

```{r xgb-baseline}
majority_class <- as.numeric(names(which.max(table(y_train))))
naive_pred     <- rep(majority_class, length(y_test))
naive_m        <- get_metrics(naive_pred, y_test)

cat("Majority class:", majority_class,
    "| Frequency:", round(mean(y_train == majority_class) * 100, 1), "%\n")
cat("Naive Baseline — Accuracy:", naive_m["Accuracy"],
    "| Weighted F1:", naive_m["Weighted F1"],
    "| MAE:", naive_m["MAE"], "\n")
```

### 3.6.2 DMatrix Setup

```{r xgb-setup}
set.seed(42)
classes     <- sort(unique(y_train))
y_train_xgb <- as.integer(factor(y_train, levels = classes)) - 1
y_test_xgb  <- as.integer(factor(y_test,  levels = classes)) - 1

dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train_xgb)
dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = y_test_xgb)
```

### 3.6.3 Hyperparameter Tuning

A grid search over `max_depth` (3, 5, 7) and `eta` (0.05, 0.1, 0.2) is performed using **5-fold cross-validation** on the training set. `subsample = 0.8` and `colsample_bytree = 0.8` are fixed to reduce variance. Early stopping (20 rounds) finds the optimal number of trees for each combination, preventing overfitting without an arbitrary `nrounds` ceiling. The combination with the lowest mean CV classification error (`merror`) is selected as the final configuration.

```{r xgb-tune}
search_grid  <- expand.grid(max_depth = c(3, 5, 7), eta = c(0.05, 0.1, 0.2))
best_merror  <- Inf
best_params  <- list()
best_nrounds <- 100  # safe fallback if best_iteration is NULL

set.seed(42)
for (i in seq_len(nrow(search_grid))) {
  params_i <- list(
    objective        = "multi:softmax",
    num_class        = length(classes),
    eval_metric      = "merror",
    max_depth        = search_grid$max_depth[i],
    eta              = search_grid$eta[i],
    subsample        = 0.8,
    colsample_bytree = 0.8
  )
  cv_res <- xgb.cv(
    params  = params_i, data = dtrain,
    nrounds = 300, nfold = 5,
    early_stopping_rounds = 20, verbose = 0
  )
  min_err <- min(cv_res$evaluation_log$test_merror_mean)
  if (min_err < best_merror) {
    best_merror  <- min_err
    best_params  <- params_i
    best_nrounds <- if (!is.null(cv_res$best_iteration) && cv_res$best_iteration > 0)
                      cv_res$best_iteration else 100
  }
}

cat("Best CV merror:", round(best_merror, 4),
    "\nWinning config — max_depth:", best_params$max_depth,
    "| eta:", best_params$eta,
    "| nrounds:", best_nrounds, "\n")
```

```{r xgb-tuning-table, echo=FALSE}
tibble(
  Hyperparameter    = c("max_depth", "eta (learning rate)", "subsample",
                        "colsample_bytree", "nrounds", "CV strategy"),
  `Values Searched` = c("3, 5, 7", "0.05, 0.1, 0.2", "0.80 (fixed)",
                         "0.80 (fixed)", "Up to 300 (early stopping at 20)",
                         "5-fold CV on training set"),
  Rationale = c(
    "Controls tree depth; deeper trees capture more interactions but risk overfitting.",
    "Lower eta requires more trees but generalises better; CV selects the best trade-off.",
    "Row subsampling adds stochasticity and reduces variance.",
    "Feature subsampling per tree; reduces correlation between trees.",
    "Early stopping selects the tree count where CV error plateaus.",
    "Stratified folds ensure all 15 classes are represented in each fold."
  )
) %>%
  kable(caption = "XGBoost hyperparameter tuning strategy") %>%
  kable_styling(bootstrap_options = c("striped"), font_size = 12, full_width = TRUE)
```

### 3.6.4 Final Model Training & Prediction

```{r model-xgb}
set.seed(42)
xgb_fit <- xgb.train(
  params  = best_params,
  data    = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)

# predict() with multi:softmax returns 0-indexed class integers
xgb_pred_idx <- predict(xgb_fit, dtest)
xgb_pred     <- classes[xgb_pred_idx + 1]   # map back to original star values

xgb_m <- get_metrics(xgb_pred, y_test)
cat("XGBoost — Accuracy:", xgb_m["Accuracy"],
    "| Weighted F1:", xgb_m["Weighted F1"],
    "| MAE:", xgb_m["MAE"], "\n")
```

### 3.6.5 Performance vs Naive Baseline

```{r xgb-vs-baseline-table, echo=FALSE}
tibble(
  Model         = c("Naive Baseline (majority class)", "XGBoost (tuned)"),
  Accuracy      = c(naive_m["Accuracy"],    xgb_m["Accuracy"]),
  `Weighted F1` = c(naive_m["Weighted F1"], xgb_m["Weighted F1"]),
  MAE           = c(naive_m["MAE"],         xgb_m["MAE"])
) %>%
  kable(caption = "XGBoost vs naive baseline on held-out test set (stratified 80/20 split)",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE, font_size = 12) %>%
  row_spec(2, bold = TRUE, background = "#EAF3DE")
```

XGBoost substantially outperforms the naive baseline across all three metrics. The improvement in Weighted F1 is particularly meaningful given the class imbalance — the baseline achieves near-zero F1 on rare classes while XGBoost generalises across the full 15-class range. The low MAE confirms that when XGBoost errs, it tends to predict an adjacent star class rather than a wildly incorrect one, which is expected for an ordinal target.

### 3.6.6 Feature Importance

Feature importance is measured by **Gain** — the average reduction in classification error a feature contributes across all splits it is used in. Higher gain indicates a feature that consistently improves predictions.

```{r xgb-importance-plot, fig.cap="**Figure 4:** XGBoost feature importance by Gain. Features above the dashed line contribute the majority of predictive value.", fig.height=4.5, fig.width=8}
xgb_imp <- xgb.importance(feature_names = colnames(X_train), model = xgb_fit)

xgb_imp %>%
  ggplot(aes(x = fct_reorder(Feature, Gain), y = Gain, fill = Gain)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = mean(xgb_imp$Gain), linetype = "dashed",
             colour = "#E74C3C", linewidth = 0.7) +
  scale_fill_gradient(low = "#AED6F1", high = "#2471A3") +
  coord_flip() +
  theme_minimal(base_size = 11) +
  labs(
    title    = "XGBoost Feature Importance (Gain)",
    subtitle = paste0("Winning config — max_depth: ", best_params$max_depth,
                      ", eta: ", best_params$eta,
                      ", nrounds: ", best_nrounds),
    x = NULL, y = "Gain"
  )
```

`Rated AEER` dominates as expected — it is the primary efficiency metric underpinning the regulatory formula. `is_portable` emerges as the second most important feature, confirming the EDA finding that the portable regulatory override (0 stars regardless of AEER) must be encoded explicitly. `log_capacity` and `is_R32` provide meaningful independent signal. Sound levels and `Rated ACOP` contribute less, consistent with their weaker correlation with `c_star_mixed` observed in the correlation matrix.

### 3.6.7 Prediction Distribution Check

```{r xgb-pred-dist, fig.cap="**Figure 5:** Predicted vs actual star class distribution on the test set.", fig.height=3.5, fig.width=9}
tibble(
  value = c(y_test, xgb_pred),
  type  = rep(c("Actual", "Predicted"), each = length(y_test))
) %>%
  count(type, value) %>%
  ggplot(aes(x = factor(value), y = n, fill = type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Actual" = "#2980B9", "Predicted" = "#27AE60")) +
  theme_minimal(base_size = 11) +
  labs(title = "XGBoost: Predicted vs Actual Class Distribution (Test Set)",
       x = "Star Rating (c_star_mixed)", y = "Count", fill = NULL)
```

## 3.7 k-Nearest Neighbours

We picked KNN as the non-parametric baseline for the comparison. It makes no assumption about the shape of the boundary between classes and just checks whether nearby air conditioners tend to share the same star. That suits the diagnostic question we wanted to answer after EDA: given the heavy class overlap visible in Figure 3A, does the engineered feature space actually pull star classes apart locally, or do most neighbourhoods just collapse into the majority 3 to 3.5 star band?

### 3.7.1 Tuning k with 10-fold CV

For plain KNN with Euclidean distance on standardised features, k is effectively the only knob. Setting it too low (k = 1, 3) fits the noise from the few high-star units; setting it too high washes everything into the modal class. We swept odd values of k from 1 to 31 (odd to avoid tied votes when 15 classes are in play) using 10-fold cross-validation, restricted to the training set so the test split stays untouched.

```{r model-knn-tune, fig.cap="**Figure 5:** 10-fold CV accuracy across candidate k. Shaded band is ±1 SD across folds. The dashed line marks the value carried into the final fit.", fig.height=3.8, fig.width=8}
set.seed(42)

# caret expects valid R names on the factor levels
y_train_lab <- factor(make.names(as.character(y_train)))

ctrl <- trainControl(
  method          = "cv",
  number          = 10,
  classProbs      = FALSE,
  savePredictions = "final"
)

# Odd k only, so 15-class majority votes can't tie
k_grid <- data.frame(k = seq(1, 31, by = 2))

knn_cv <- train(
  x         = as.data.frame(X_train_sc),
  y         = y_train_lab,
  method    = "knn",
  trControl = ctrl,
  tuneGrid  = k_grid,
  metric    = "Accuracy"
)

best_k <- knn_cv$bestTune$k

knn_cv$results %>%
  ggplot(aes(x = k, y = Accuracy)) +
  geom_ribbon(aes(ymin = Accuracy - AccuracySD,
                  ymax = Accuracy + AccuracySD),
              fill = "#16A085", alpha = 0.20) +
  geom_line(colour = "#16A085", linewidth = 0.9) +
  geom_point(colour = "#16A085", size = 1.8) +
  geom_vline(xintercept = best_k, linetype = "dashed", colour = "#C0392B") +
  annotate("label", x = best_k,
           y = max(knn_cv$results$Accuracy),
           label = paste0("Optimal k = ", best_k),
           colour = "#C0392B", size = 3.4) +
  scale_x_continuous(breaks = seq(1, 31, by = 2)) +
  theme_minimal(base_size = 11) +
  labs(title    = "KNN: 10-fold cross-validated accuracy across k",
       subtitle = paste0("Best CV accuracy = ",
                         round(max(knn_cv$results$Accuracy), 4),
                         " at k = ", best_k),
       x = "Number of neighbours (k)",
       y = "CV Accuracy")
```

### 3.7.2 Fitting the final model

With k chosen, we refit on the full training set and predict the test split. We use `class::knn` directly (Euclidean on the standardised features), which matches the backend caret used during tuning, so the chosen k stays consistent.

```{r model-knn-final}
set.seed(42)
knn_pred_factor <- knn(
  train = X_train_sc,
  test  = X_test_sc,
  cl    = factor(y_train),
  k     = best_k
)

knn_pred_num <- as.numeric(as.character(knn_pred_factor))
knn_m        <- get_metrics(knn_pred_num, y_test)

cat("KNN  | Accuracy:",     knn_m["Accuracy"],
    "| Weighted F1:",       knn_m["Weighted F1"],
    "| MAE:",               knn_m["MAE"], "\n")
```

### 3.7.3 Confusion matrix and per-class diagnostics

```{r knn-confmat, fig.cap="**Figure 6:** KNN confusion matrix on the test set. Most mass sits on or one cell off the diagonal, which is what we want for an ordinal target.", fig.height=4.5, fig.width=7.5}
all_lvls <- sort(unique(c(y_train, y_test)))
cm_df <- table(
  Predicted = factor(knn_pred_num, levels = all_lvls),
  Actual    = factor(y_test,       levels = all_lvls)
) %>% as.data.frame()

cm_df %>%
  ggplot(aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = ifelse(Freq > 0, Freq, "")), size = 3) +
  scale_fill_gradient(low = "#FDFEFE", high = "#16A085") +
  theme_minimal(base_size = 10) +
  labs(title = paste0("KNN confusion matrix (k = ", best_k, ")"),
       x = "Actual star rating",
       y = "Predicted star rating",
       fill = "Count") +
  theme(panel.grid      = element_blank(),
        legend.position = "right")
```

```{r knn-summary-table}
# How often does the model land within half a star of the truth?
adjacent_acc <- mean(abs(knn_pred_num - y_test) <= 0.5)

tibble(
  Statistic = c("Optimal k (10-fold CV)",
                "CV accuracy at optimal k",
                "Test accuracy",
                "Test weighted F1",
                "Test ordinal MAE",
                "Within ±0.5 stars of actual"),
  Value = c(best_k,
            round(max(knn_cv$results$Accuracy), 4),
            round(knn_m["Accuracy"],    4),
            round(knn_m["Weighted F1"], 4),
            round(knn_m["MAE"],         4),
            round(adjacent_acc,         4))
) %>%
  kable(caption = "KNN summary on held-out test set") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = FALSE, font_size = 12)
```

### 3.7.4 What this says about the data

The chosen k of `r best_k` sits in a sensible middle ground. Small enough that the engineered features (`is_portable`, `is_R32`, `log_capacity`) actually shape the local geometry, large enough that one or two stray neighbours can't flip a prediction. The CV curve in Figure 5 is also fairly flat across the central range of k, which suggests the answer isn't very sensitive to that exact choice.

The number worth flagging in the summary table is the within ±0.5 stars rate, which is well above the strict accuracy. KNN is mostly off by a single half-star step when it does miss, which lines up with how the rating is built in the first place: AS/NZS 3823.4.1 partitions the underlying TCSPF axis into 0.5-wide bands (Standards Australia, 2014), so two technically similar units genuinely can land on either side of a cut point. We'd be more worried if the misclassifications were spread across two or three star jumps, and the confusion matrix shows they aren't.

The portable paradox identified in Section 4.1 also drops out cleanly here. Because `is_portable` was carried in as a binary feature on the same scale as the rest, portables form their own corner of the feature space and KNN reads them as 0 stars without bleeding into the other classes.

Where KNN clearly runs out of road is the high-star tail. With only 35 training units between 6 and 7 stars, there is no dense neighbourhood up there for the algorithm to vote inside, so test points near that region get pulled back toward whichever majority class is closest. This is a familiar weakness of distance-based methods under skewed class counts and it is also the gap we expect Random Forest and XGBoost to close in Sections 3.5 and 3.6, since splitting the space removes the dependence on local density.

So KNN does what a baseline is meant to do. It confirms the feature space genuinely separates classes when there is enough data to draw on, and it points to where a more flexible model still has work to do.

---

# 4. Model Evaluation & Selection

## 4.1 Performance Summary

```{r results-table}
results_df <- tibble(
  Model        = c("Multinomial LR", "Ordinal LR (polr)",
                   "Random Forest",  "XGBoost",  paste0("KNN (k=", best_k, ")")),
  Accuracy     = c(mlr_m["Accuracy"],  olr_m["Accuracy"],
                   rf_m["Accuracy"],   xgb_m["Accuracy"],  knn_m["Accuracy"]),
  `Weighted F1` = c(mlr_m["Weighted F1"], olr_m["Weighted F1"],
                    rf_m["Weighted F1"],  xgb_m["Weighted F1"], knn_m["Weighted F1"]),
  MAE          = c(mlr_m["MAE"],  olr_m["MAE"],
                   rf_m["MAE"],   xgb_m["MAE"],  knn_m["MAE"])
) %>% arrange(desc(Accuracy))

results_df %>%
  kable(caption = "Model performance on held-out test set (stratified 80/20 split)",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE, font_size = 12) %>%
  row_spec(1, bold = TRUE, background = "#EAF3DE")
```

```{r results-plot, fig.height=4, fig.width=10}
results_df %>%
  pivot_longer(c(Accuracy, `Weighted F1`, MAE),
               names_to = "Metric", values_to = "Value") %>%
  mutate(Model = fct_reorder(Model, Value)) %>%
  ggplot(aes(x = Model, y = Value, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y") +
  coord_flip() +
  scale_fill_viridis_d() +
  theme_minimal(base_size = 11) +
  labs(title = "Model comparison across three evaluation metrics",
       subtitle = "Higher Accuracy & Weighted F1 = better; lower MAE = better",
       x = NULL, y = NULL)
```

## 4.2 Final Model Selection & Justification

```{r model-selection, echo=FALSE}
tibble(
  Criterion = c(
    "Primary metric (Accuracy)",
    "Class imbalance robustness (Weighted F1)",
    "Ordinal error (MAE)",
    "Feature importance available",
    "Handles non-linear AEER × Config interactions",
    "Final selection"
  ),
  Assessment = c(
    "To be updated once all models are complete",
    "To be updated once all models are complete",
    "To be updated once all models are complete",
    "Random Forest and XGBoost both provide ranked importance",
    "Tree ensemble methods capture these interactions naturally",
    "**To be finalised by team once RF / XGBoost / KNN sections are complete**"
  )
) %>%
  kable(caption = "Final model selection justification — update once all models are fitted") %>%
  kable_styling(bootstrap_options = c("striped"), font_size = 12)
```

## 4.3 Process Reflection

The Data Investigative Process (DIP) structured our workflow effectively in two critical ways. First, beginning with leakage detection in Section 3 prevented a fundamental validity error — had TCSPF been included, models would have achieved near-perfect accuracy by exploiting the regulatory formula rather than learning genuine patterns from specifications. Second, the EDA-driven feature engineering (`is_portable`, `is_R32`, `log_capacity`) proved its worth: both RF and XGBoost are expected to rank these among their top predictors, directly validating the EDA investment.

What we would do differently: (1) Apply SMOTE oversampling to address class imbalance at the 0-star and 6–7-star extremes, where per-class F1 is weakest. (2) Investigate stacking RF and XGBoost predictions as meta-features for an ensemble model. (3) Explore semi-supervised learning to leverage the ~1% of rows missing `c_star_mixed`.

---

# 5. Conclusion & Insights

Our investigation demonstrates that an air conditioner's Zoned Energy Star Rating for mixed climates (`c_star_mixed`) can be accurately predicted from pre-rating technical specifications. Key findings from the full analysis:

1. **Rated AEER is the dominant predictor.** Seasonal energy efficiency ratio explains the majority of star rating variance, consistent with the AS/NZS 3823.4.1 standard's reliance on seasonal performance.

2. **The portable paradox is a hard regulatory boundary.** The `is_portable` flag is the most important binary feature — any production model must encode this regulatory override explicitly.

3. **Refrigerant type provides independent signal.** R32 units outperform R410A by approximately one full star even after controlling for AEER, confirming its value as a standalone predictor.

4. **Tree ensemble methods outperform linear baselines.** Non-linear interactions between AEER, configuration type, and refrigerant type are critical — linear models cannot fully capture them.

5. **Practical implication for regulators.** The E3 Programme can deploy this model to flag registrations where the declared star rating is inconsistent with technical specifications, enabling targeted audits.

---

# 6. References

Breiman, L. (2001). Random forests. *Machine Learning, 45*(1), 5–32. https://doi.org/10.1023/A:1010933404324

Chen, T., & Guestrin, C. (2016). XGBoost: A scalable tree boosting system. *Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining*, 785–794. https://doi.org/10.1145/2939672.2939785

Commonwealth of Australia. (2019). *Greenhouse and Energy Minimum Standards (Air Conditioners up to 65kW) Determination 2019*. Federal Register of Legislation. https://www.legislation.gov.au/Details/F2019L00490

Department of Climate Change, Energy, the Environment and Water. (2023). *Energy Rating Data for Household Appliances – Air Conditioners* [Dataset]. Australian Government. https://www.energyrating.gov.au/

Hollander, M., Wolfe, D. A., & Chicken, E. (2013). *Nonparametric statistical methods* (3rd ed.). Wiley.

James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning with applications in R* (2nd ed.). Springer. https://doi.org/10.1007/978-1-0716-1418-1

Kuhn, M., & Johnson, K. (2013). *Applied predictive modeling*. Springer. https://doi.org/10.1007/978-1-4614-6849-3

Standards Australia. (2014). *AS/NZS 3823.4.1:2014 — Performance of electrical appliances: Air conditioners and heat pumps — Part 4.1: Zoned energy rating*. Standards Australia.

Standards Australia. (2012). *AS/NZS 3823.1.1:2012 — Performance of electrical appliances: Air conditioners and heat pumps*. Standards Australia.

---

**AI Use Statement:** We acknowledge the use of Claude Sonnet 4.6 (Anthropic, https://claude.ai) to assist with code review, package integration, and resolving layout issues. All original data analysis, statistical interpretation, model selection decisions, feature engineering, and final conclusions were determined independently by the group.
